Satellite-detected drought frequencies predict plant life history strategies

### 2018-11-26 manuscript/manuscript.tex
round 2

\documentclass[man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Plant life history strategies predicted by satellite-detected drought},
            pdfauthor={J Grey Monroe, Brian Gill, \& Kathryn Turner},
            pdfkeywords={drought adaptation, life history evolution, remote sensing,
phylogeography, herbaria records},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Plant life history strategies predicted by satellite-detected drought}
    \author{J Grey Monroe\textsuperscript{1,2}, Brian Gill\textsuperscript{3}, \&
Kathryn Turner\textsuperscript{4}}
    \date{}

\shorttitle{Drought and life history}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO\\\textsuperscript{2} College of Agriculture, Colorado State University, Fort Collins, CO\\\textsuperscript{3} Department of Ecology and Evolutionary Biology, Brown University, Provience, RI\\\textsuperscript{4} Biology Department, Pennsylvania State University, State College, PA}
\keywords{drought adaptation, life history evolution, remote sensing, phylogeography, herbaria records\newline\indent Word count: }
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\linenumbers

@@

Correspondence concerning this article should be addressed to J Grey
Monroe, 307 University Ave, Fort Collins, CO 80523. E-mail:
\href{mailto:monroejg@colostate.edu}{\nolinkurl{monroejg@colostate.edu}}}

\abstract{
Identifying the environmental factors that predict the evolution of
annual and perennial life history strategies in plants is important for
understanding ecosystem functioning, perennial cropping systems, and is
a long-standing goal of evolutionary ecology. A classic hypothesis is
that annual and perennial strategies reflect adaptation to environments
that differ in relative drought frequency. We test this hypothesis in
Heliophila (Brassicaceae), a genus of flowering plants native to
Southern Africa using herbarium occurrence records and
satellite-detected drought histories. We find that perennial Heliophila
species are observed in environments where droughts are significantly
less frequent compared to annuals, especially with regards to summer
drought. These correlations remain predictive while controlling for
phylogeny, lending support to the hypothesis that drought associated
natural selection has shaped differences in the distributions of these
strategies. Additionally, the collection dates of herbarium records are
consistent with a scenario in which annual species escape drought prone
seasons. Together these results point to a model of evolution that
supports classical hypotheses favoring perenniality in environments that
experience less drought, compared to annuals which escape drought prone
seasons as seed. Thus, changing drought regimens during the Anthropocene
may threaten locally adapted species. 
}

@@

\hypertarget{life-history}{%
\subsection{Life history}\label{life-history}}

Plants exhibit extraordinary diversity in life history strategies, from
herbaceous species that complete the entire seed to seed life cycle in a
number of weeks \{Guo, 2012 \#73\} to trees that live for thousands of
years \{Brown 1996 \url{http://www.rmtrr.org/oldlist.htm} \}. Along this
continuum an important division exists, distinguishing annuals which
complete their seed to seed life cycle within a single calendar year
from perennials which can persist over multiple years and therefore
experience the entire range of seasonal conditions. The ecological
factors that explain the evolution of these alternative strategies
remains poorly resolved \{Friedman, 2015 \#523\}. Understanding the
drivers of selection for these alternative strategies is important
because annuals and perennials have differing impacts on ecosystem
functioning, such as higher nitrogen concentration in perennials and
larger specific root length in annuals that affect nutrient cycling
\{Garnier, 1997 \#280\}\{Roumet, 2006 \#738\}. Furthermore, predicting
the conditions that favor these strategies in nature is useful for
developing more sustainable perennial cropping systems \{Lelievre and
Volaire, 2009\}.

\hypertarget{hypotheses}{%
\subsection{Hypotheses}\label{hypotheses}}

Seasonal drought conditions can represent a severe challenge to the
persistence of individual organisms. Well-established hypotheses suggest
that an annual life history is adaptive in environments where frequent
droughts makes an escape from stressful drought conditions in the form
of seed advantageous \{Stearns 1992\}\{Silvertown and Charlesworth,
2001\}. These hypotheses are based on models of life history evolution
in which short life cycles are favored in unpredictable environments or
those with frequent stressful conditions. Plants are expected to be more
likely to survive a drought that cooccurs with the seed phase of a
plant's life cycle in which they are most resistant to desiccation.
Previous efforts to address this hypothesis have yielded mixed or
qualitative results. Transitions to annuality in Oenothera were
associated with warmer summers and drier winters, but not with increased
drought directly \{Evans, 2005 \#260\}. Annual and perennial species of
Nemesia were qualitatively associated with winter rather and summer
rainfall environments respectively \{Datson, 2008\}. Similarly, annual
species of Scorzoneroides were associated with environments classified
as \enquote{unpredictable} \{Cruz-Mazo, 2009 \#598\}. Annual and
perennial varieties of the wild rice species Oryza rufipogon were
observed more frequently in dry and humid climates, respectively
\{Morishima et al.~1984\}. While these reports suggest the prominence of
annuals in environments considered more arid, a quantitative assessment
of how historic drought frequency may favor annual or perennial life
history strategies remains lacking.

Here we combine a long-term dataset of global imaging with metadata from
natural history collections to test classic hypotheses about the
evolution of life history strategies within the African endemic mustard
genus, Heliophila L. (Brassicaceae). If annual species temporally escape
drought as seed, then drought frequency should be an important
determinant of the distribution of life history strategies across the
landscape, and annual species should be more commonly associated with
drought prone regions than perennial species. Furthermore, if annual
species have adapted to escape drought prone seasons as seeds,
observations of annual species should be rare during drought prone
seasons. Phylogenetic relatedness can have significant non-random
effects on species distributions and life history traits (cite), and
therefore we assessed the relationship between life history distribution
and drought frequency in a phylogenetically controlled background.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

@@

\hypertarget{drought}{%
\subsubsection{Drought}\label{drought}}

Remotely sensed data is a valuable tool for characterizing seasonal
patterns in drought \{AghaKouchak 2015\}. With these data, we can
exploit the fact that satellite imagery can detect reductions in plant
cover and moisture at landscape scales that are characteristic of
drought conditions. Unlike weather station data, which is limited in
geographic coverage, and manual observation, which is limited in scope,
satellite detected data provides a long term global historical record of
drought events at fine temporal and spatial resolutions. One such
measure, the Vegetative Health Index (VHI), has been collected since
1981 by NOAA AVHRR satellites and is a combined measure of thermal and
vegetative parameters. VHI quantifies drought induced vegetative stress
weekly by detecting reduced plant cover based on the normalized
difference vegetation index (Vegetative Condition Index) combined with
reductions in moisture associated with anomalies in thermal spectra
(Temperature Condition Index). Put most simply, the VHI detects drought
by identifying environments that appear unusually brown and dry for that
location and time of the year.

This data provides a global, long term, and quantitative perspective on
drought variability. The VHI database (see methods) presents a valuable
resource to study seasonal patterns in the frequency of drought across
environments and to test hypotheses about the effect of drought on
ecological and evolutionary processes. As such, it has been validated as
a tool for detecting drought and predicting crop yields \{Kogan 1997\}.
It has also proven useful for predicting infraspecific variation in
drought tolerance traits and genes \{Mojica et al 2016\}\{Dittberner et
al 2018\}\{Monroe et al 2018\}.

VHI quantifies drought induced vegetative stress weekly by detecting
reduced plant cover based on the normalized difference vegetation index
(Vegetative Condition Index) combined with reductions in moisture
associated with anomalies in thermal spectra (Temperature Condition
Index). Put most simply, the VHI detects drought by identifying
environments that appear unusually brown and dry for that location and
time of the year.

The Vegetative Health Index (VHI) was used to examine seasonal drought
frequency at the collection locations of Heliophila herbarium specimens.
The VHI is a satellite-detected drought measurement method based on
observations of vegetative stress caused by drought \{Kogan, 1995
\#534\}, combining deviations from historic climatic (Temperature
Condition Index) and vegetative conditions (Vegetative Condition Index).
The VHI has been measured weekly at 16km2 spatial resolution since XXXX.
The frequencies of observing drought conditions (VHI\textless{}40,
standard recommended by NOAA developers of VHI) during the southern
hemisphere spring (quarter surrounding spring equinox), summer (quarter
surrounding summer solstice), fall (quarter surrounding fall equinox),
and winter (quarter surrounding winter solstice) were calculated
globally from 1981 to 2015. These values were calculated for the
collection location of each Heliophila herbarium specimen in our
filtered dataset.

NDVI = (Ch2 - Chl)/(Ch2 + Chl)

\[NDVI = \frac{NIR - Red}{NIR + Red}\]

\[VCI = 100\frac{NDVI - NDVI_{min}}{NDVI_{max} - NDVI_{min}}\]

\[TCI = 100\frac{T - T{min}}{T{max} - T{min}}\]

\[VHI = 0.5(VCI) + 0.5(TCI)\] This data provides a global, long term,
and quantitative perspective on drought variability. The VHI database
presents a valuable resource to study seasonal patterns in the frequency
of drought across environments and to test hypotheses about the effect
of drought on ecological and evolutionary processes. As such, it has
been validated as a tool for detecting drought and predicting crop
yields ({\textbf{???}}). It has also proven useful for predicting
infraspecific variation in drought tolerance traits and genes
(Dittberner et al., 2018; Mojica et al., 2016; Monroe et al., 2018).

\hypertarget{life-history-1}{%
\subsubsection{Life history}\label{life-history-1}}

Heliophila is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions which are among the most botanically diverse environments on
Earth. Indeed, the estimated \textasciitilde{}50 Heliophila species are
considered to represent the most diverse genus of the family
Brassicaceae \{Mummenhoff, 2005 \#494\}. This genus includes both
perennial and annual species and this change in life history strategy
has likely arisen multiple independent times \{Appel \& Al-Shehbaz,
1997\}\{Mummenhoff, 2005 \#494\}. These multiple transitions between
life history strategy and it's well documented record in global herbaria
provided the ideal system to address classic hypotheses about the
evolutionary drivers of annual and perennial life history strategies.

\hypertarget{phyologeny}{%
\subsubsection{Phyologeny}\label{phyologeny}}

Heliophila is a charismatic genus of flowering plants from the
Brassicaceae provides a valuable model to study the evolution of annual
and perennial life history strategies because each has arisen
independently multiple times within the genus. These independent origins
allow for phylogenetically constrained analyses comparing the climate
distributions of annual and perennial species, to gain greater insight
into the role of selection.

Aligned Heliophila ITS sequences were obtained from previous work by
Mandáková et al.~\{Mandáková, 2012 \#339\}. Aethionema, Alliaria,
Cardamine, Chamira, and Rorippa ITS records from were downloaded from
Genbank (link) and aligned together with Heliophila ITS sequences using
MAFFT as outgroups? (cite). Model selection for construction of
phylogeny was performed in jModeltest2 with CIPRES (cite). Based on this
analysis, \(GTR + L\) were selected. Ultrametric phylogeny was estimated
with branch lengths as relative time (details). Life history character
states were used based on habit reported by Mummemhoff et
al.~\{Mummenhoff, 2005 \#494\}. Species reported to have any form of
perennial life history (such as biennial\ldots{}) were classified for
analyses here as perennial, where species reported to have strictly
annual life histories were classified for analyses here as annual.

\hypertarget{herbarium-specimens}{%
\subsubsection{Herbarium Specimens}\label{herbarium-specimens}}

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years (Lang et al 2018). These
collections, housed in herbaria, comprise an enormous, yet largely
untapped, ecological dataset. These specimens are increasingly being
recognized as an invaluable and underutilized source of data pertaining
to biological responses to past environmental conditions (Pyke et
al.~2010). Herbarium specimens and their associated metadata have been
used since the 1960s to study species' geographical distributions
(reviewed in Lang et al.~2018). Herbarium specimens have been used to
track plant responses to climate change using herbarium time-series,
including the relationships between traits, geography, and climate.
\{Wolf et al.~2016, Davis et al.~2015, Stropp et al.~2017\}. A
potentially fruitful approach combines herbarium data with remotely
sensed data sets that characterize the environmental and climatic
conditions across landscapes. For example, remotely sensed global
ultraviolet-B radiation exposure may explain differences in trichome
phenotypes, observed from herbarium specimens, between the native and
invasive ranges of two herbaceous plants (Václavík et al.~2017).

Herbarium specimen records for all Heliophila species were downloaded
from the Global Biodiversity Information Facility (gbif.org) on July 21,
2018. Several filtering steps were employed to clean this data (see Sup
Mat. R code). For these analyses, records were restricted to those which
are associated with a physical herbarium specimen, rather than other
types of observations. Consequentially, this sample represents
individuals in a narrow range of vegetative and/or flowering states --
for example, herbarium specimens cannot be collected for individuals
existing as seed. Analyses were restricted to Heliophila species with
previously-reported life history information \{Mummenhoff, 2005 \#494\};
42 of \textasciitilde{}50 named species in this genus were used. Only
records including geo-referenced information (latitude and longitude of
collection site) were used for downstream analyses of historic drought
frequency. To discard records likely to be erroneous and to restrict
analyses to plants in their native, non-cultivated, habitats, records
occurring outside of the native range (roughly, the southern half of
Africa) were removed. This excluded specimens such as those growing in
likely cultivation in Australia or terrestrial plants recorded from
non-terrestrial oceanic locations.. Finally, duplicate records, based on
species, collection location coordinates, and date, were removed.

We downloaded 8670 records from GBIF.

@@

\hypertarget{herbarium-clean-up}{%
\subsubsection{Herbarium clean up}\label{herbarium-clean-up}}

We filtered raw GBIF by restricting our analyses to

\begin{itemize}
\tightlist
\item
  Species with reported life history\\
\item
  Records with geospatial data\\
\item
  Records from collection sites classified as land pixels
\item
  Records for perserved specimens
\item
  Records from Africa
\item
  Non-duplicate records (ie. identical species, location, collection
  date)
\end{itemize}

\hypertarget{phylogeny}{%
\subsubsection{Phylogeny}\label{phylogeny}}

\hypertarget{contrast}{%
\subsubsection{Contrast}\label{contrast}}

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial Heliophila species would be significantly associated with
historic drought frequency. To do so, the average drought frequency for
each season was calculated across the herbarium specimen collection
locations for each species. The relationship between drought frequencies
across each taxon's range and life habitat (annual or perennial) was
evaluated using Firth's penalized-likelihood logistic regression and
phylogenetic logistic regression.

@@

Out of 8670 \emph{Heliophila} GBIF recrods, 6634 were for species with
reported life history (Mummenhoff, Al-Shehbaz, Bakker, Linder, \&
Mühlhausen, 2005), 3653 had geospatial data, 3460 were located on pixels
classified as land having drought measurements, 3457 were located in
Africa, 3162 were not duplicated. After all filtering steps, 2192
records for 42 species (Figure \ref{tab:gbifrecordsmap}, Table
\ref{tab:speciesmeanstable}) passed.

\begin{figure}
\centering
\includegraphics{../figures/gbifrecords.pdf}
\caption{\label{fig:gbifrecordsmap}Map of 2192 GBIF records that passed quality
filtering.}
\end{figure}

\hypertarget{drought-frequency.}{%
\subsubsection{Drought frequency.}\label{drought-frequency.}}

\begin{figure}
\centering
\includegraphics{../figures/life_history_drought_boxplots.pdf}
\caption{\label{fig:boxplots}Comparison of drought frequency across seasons measured
at the GBIF records of annual and perennial species of Heliophila.}
\end{figure}

>>>>

\documentclass[man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Satellite-detected drought frequencies predict plant life history strategies},
            pdfauthor={J. Grey Monroe, Brian Gill, Kathryn Turner, \& John K. McKay},
            pdfkeywords={drought adaptation, life history evolution, remote sensing,
phylogeography, herbaria records},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Satellite-detected drought frequencies predict plant life history
strategies}
    \author{J. Grey Monroe\textsuperscript{1,2}, Brian Gill\textsuperscript{3},
Kathryn Turner\textsuperscript{4}, \& John K. McKay\textsuperscript{2}}
    \date{}

\shorttitle{Drought and life history}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO 80523, USA\\\textsuperscript{2} College of Agriculture, Colorado State University, Fort Collins, CO 80523, USA\\\textsuperscript{3} Institute for Environment and Society, Brown University, Providence, RI 02912, USA\\\textsuperscript{4} Biology Department, Pennsylvania State University, State College, PA 16802, USA}
\keywords{drought adaptation, life history evolution, remote sensing, phylogeography, herbaria records}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\linenumbers
\newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}

@@

Correspondence concerning this article should be addressed to J. Grey
Monroe, 307 University Ave, Fort Collins, CO 80523. E-mail:
\href{mailto:monroejg@colostate.edu}{\nolinkurl{monroejg@colostate.edu}}}

\abstract{
Explaining variation in life history strategies is a long-standing goal
of evolutionary biology. For plants, annual and perennial life histories
are thought to reflect adaptation to environments that differ in the
frequency of environmental stress such as drought. Here we test this
hypothesis in \textit{Heliophila} (Brassicaceae), a diverse genus of
flowering plants native to Southern Africa using herbarium occurrence
records and satellite-detected drought histories. We find that perennial
\textit{Heliophila} species are found in environments where droughts are
significantly less frequent compared to annuals. These associations are
predictive while controlling for phylogeny, lending support to the
hypothesis that drought related natural selection has influenced the
distributions of these strategies. Additionally, the difference in
drought frequency between annual and perennial species distributions is
greatest during the summer and fall, which also appears to be when
annuals are in the seed phase of their life cycle based on collection
dates of annual species. Together, these findings support traditional
hypotheses about the drivers of life history strategy in plants - that
perenniality is favored in environments with less frequent drought and
annuality in enviroments with more frequent drought by allowing them to
escape drought prone seasons as seeds. 
}

@@

\hypertarget{life-history-variation-in-plants}{%
\subsection{Life history variation in
plants}\label{life-history-variation-in-plants}}

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals (i.e.~monocarpic or semelparous)
which complete their seed to seed life cycle within a single calendar
year from perennials (i.e.~polycarpic or iteroparous{]} which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed. In contrast, perennial plants can continue vegetative growth after
reproduction and experience all seasons. These represent fundamentally
different life history strategies, but the ecological factors that
explain their evolution remain uresolved (Friedman \& Rubin, 2015).

Identifying the drivers of selection for these alternative strategies is
important because annuals and perennials have differing impacts on
ecosystem functioning. For example, perennials have higher nitrogen
concentration and larger specific root length in annuals, which both
affect nutrient cycling (Garnier, Cordonnier, Guillerm, \& Sonié, 1997;
Roumet, Urcelay, \& Dı'az, 2006). Furthermore, predicting the conditions
that favor these strategies in nature is useful for developing more
sustainable perennial cropping systems (Lelièvre \& Volaire, 2009).

Classical theory predict shorter life spans and annuality in
environments where adult mortality is high (Charnov \& Schaffer, 1973;
Franco \& Silvertown, 1996; Stearns, 1992). In plants, this has been
extended to the hypothessis that annuality is adaptive when it allows
plants to escape drought (Schaffer \& M, 1975). Lack of water is perhaps
the greatest threat to survival during vegetatitive or reproductive
growth and annuals can remain dormant (and protected as a seed) during
drought. Thus, environments with greater seasonal drought frequency may
select for annual life histories that complete reproduction prior to
drought prone seasons. Conversely, environments with less frequent
drought may select for perennial species, which may benefit from
multiple bouts of reproduction and competitive advantage (Corbin \&
D'Antonio, 2004). These prediction has been supported by the association
of annualilty with arid environments in wild rice (Morishima, Sano, \&
Oka, 1984) and \emph{Oenothera} (Evans, Hearn, Hahn, Spangle, \&
Venable, 2005). Additionally, annual and perennial species of Nemesia
were qualitatively associated with winter rather and summer rainfall
environments respectively (Datson, Murray, \& Steiner, 2008) and annual
species of Scorzoneroides were associated with environments classified
as unpredictable (Cruz-Mazo, Buide, Samuel, \& Narbona, 2009). These
findings provide qualitative evidence supporting the prediction that
annual species are found in environments that experience more frequent
drought, and vice versa, but whether the history frequency of drought
events indeed predicts the distributions annual or perennial life
history strategies has yet to be tested.

Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural collections to test these classic
hypotheses about the evolution of life history strategies within the
African endemic mustard genus, \emph{Heliophila} L. (Brassicaceae). If
annuallity is an adaptive strategy allowing plants to escape drought
prone seasons, then drought frequency should predict the distribution of
life history strategies across landscapes, and annual species should be
more commonly associated with drought prone regions than perennial
species. Furthermore, if annual species have adapted to escape drought
prone seasons as seeds, observations of annual species should be rare
during drought prone seasons. Phylogenetic relatedness can have
significant non-random effects on species distributions and life history
traits {[}{]}, and therefore we assessed the relationship between life
history distribution and drought frequency in a phylogenetically
controlled background.

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

\hypertarget{availability}{%
\subsection{Availability}\label{availability}}

All data and the source code to produce this manuscript are available at
\url{https://github.com/greymonroe/heliophila}.

@@

\hypertarget{satellite-detected-drought-data}{%
\subsubsection{Satellite-detected drought
data}\label{satellite-detected-drought-data}}

Remotely sensed data is a powerful tool for characterizing seasonal
patterns in drought because it is less limited in spatial and temporal
scope and resolution than weather stations or field observations
(AghaKouchak et al., 2015). To quantify the frequency of drought during
different seasons across landscapes, we used the remotely sensed
Vegetative Health Index (VHI), which measures landscape scale reductions
in plant cover and temperature conditions characteristic of drought
(Kogan, 2001). Generated from data collected by NOAA AVHRR satellites
since 1981, the VHI combines normalized difference vegetation index
(NDVI) derived measures of vegetative stress (Vegetative Condition Index
- VCI) with temperature stress indicated by anomalies in thermal spectra
(Temperature Condition Index - TCI).

The VHI of year \(y\) during week \(w\) of \([1,52]\) at pixel \(i\) is
derived from the following equations, where \(n\) is the number of years
observed.

\[VCI_{y,w,i} = 100\frac{NDVI_{y,w,i} - NDVI_{min}}{NDVI_{max} - NDVI_{min}}\]

where \(NDVI_{min} = min(NDVI_{1981,w,i}...NDVI_{1981+n,w,i})\) and
\(NDVI_{max} = max(NDVI_{1981,w,i}...NDVI_{1981+n,w,i})\)

\[TCI_{y,w,i} = 100\frac{T_{y,w,i} - T_{min}}{T_{max} - T_{min}}\]

where \(T_{min} = min(T_{1981,w,i}...T_{1981+n,w,i})\) and
\(T_{max} = max(T_{1981,w,i}...T_{1981+n,w,i})\)

\[VHI_{y,w,i} = 0.5(VCI_{y,w,i}) + 0.5(TCI_{y,w,i})\]

Thus, VHI measurements are standardized according to conditions
historically observed at each locations. These measurements have been
validated and generally used for evaluating drought risk and predicting
crop yields in agriculture (e.g., Rojas, Vrieling, \& Rembold, 2011;
Kogan et al., 2016). But they also present a tool to study seasonal
patterns in the frequency of drought across environments and to test
hypotheses about the effect of drought on ecological and evolutionary
processes. Indeed, the VHI has been applied recently to study drought
related ecology of natural species and proven useful for predicting
infraspecific variation in drought tolerance traits and genes
(Dittberner et al., 2018; Mojica et al., 2016; Monroe et al., 2018).
Here, we accessed VHI data at \(16km^2\) resolution from 1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} sepecies.

\hypertarget{life-history-data-for-heliophila}{%
\subsubsection{Life history data for
Heliophila}\label{life-history-data-for-heliophila}}

\emph{Heliophila} is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions. These are among the most botanically diverse environments on
Earth and the estimated \textasciitilde{}50 \emph{Heliophila} species
are considered to be the most diverse genus of the family Brassicaceae
(Mummenhoff, Al-Shehbaz, Bakker, Linder, \& Mühlhausen, 2005). This
genus includes both perennial and annual species and this change in life
history strategy has likely arisen multiple independent times (Appel \&
Al-Shehbaz, 1997; Mummenhoff et al., 2005). We used life histories
reported by Mummenhoff et al. (2005), grouping species with annual or
perennial life histories. Perenniality was defined based on the
traditional defition and included any form of perennial life history
(e.g., herbs, shrubs, mixed, etc).

\hypertarget{heliophila-herbarium-specimens}{%
\subsubsection{Heliophila herbarium
specimens}\label{heliophila-herbarium-specimens}}

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years. Herbarium specimens and
their associated metadata have been used since the 1960s to study
species' geographical distributions (reviewed by Willis et al. (2017)
and Lang, Willems, Scheepens, Burbano, and Bossdorf (2018)). And as they
become digitized (Soltis, 2017), these collections have been used to
study relationships between trait distributions, geography, and climate
(Davis, Willis, Connolly, Kelly, \& Ellison, 2015; Stropp et al., 2016;
Václavı'k, Beckmann, Cord, \& Bindewald, 2017; Wolf, Zimmerman,
Anderegg, Busby, \& Christensen, 2016). To characterize the diributions
of annual and perennial \emph{Heliophila} species, all records for the
genus \emph{Heliophila} were downloaded from the Global Biodiversity
Information Facility (gbif.org) on July 21, 2018.

\hypertarget{sequence-data-for-phylogeny}{%
\subsubsection{Sequence data for
phylogeny}\label{sequence-data-for-phylogeny}}

Aligned Heliophila ITS sequences were obtained from previous work by
Mandáková et al. (2012). Aethionema, Alliaria, Cardamine, Chamira, and
Rorippa ITS records from were downloaded from Genbank {[}{]} and aligned
together with Heliophila ITS sequences using MAFFT as outgroups {[}{]}.

@@

\hypertarget{drought-frequency-calculations}{%
\subsubsection{Drought frequency
calculations}\label{drought-frequency-calculations}}

To characterize drought regimens across the distrubtions of annual and
perennial species of \emph{Heliohpila}, we calculated drought during
different seasons at the location of observations for \emph{Heliophila}
records using the VHI. Specifically, we created global maps of the
frequencies of observing drought conditions (VHI\textless{}40, NOAA)
during the winter (quarter surrounding winter solstice), spring (quarter
surrounding spring equinox), summer (quarter surrounding summer
solstice) and fall (quarter surrounding fall equinox) from 1981 to 2015.
From these maps, the drought frequency during the winter, spring,
summer, and fall were extracted for the locations of all GBIF records.

\hypertarget{herbarium-record-quality-control}{%
\subsubsection{Herbarium record quality
control}\label{herbarium-record-quality-control}}

To remove instance with spurious location data, we filtered raw GBIF by
restricting our analyses to include only:\\
* Species with reported life history\\
* Records with geospatial data\\
* Records without known geospatial issues\\
* Records from collection sites classified as land pixels\\
* Records from Africa\\
* Non-duplicate records (ie. identical species, location, collection
date)

\hypertarget{phylogeny-construction}{%
\subsubsection{Phylogeny construction}\label{phylogeny-construction}}

Model selection for construction of phylogeny was performed in
jModeltest2 with CIPRES (cite). Based on this analysis, \(GTR + L\) were
selected. Ultrametric phylogeny was estimated with branch lengths as
relative time (details).

\hypertarget{comparison-of-drought-frequency-between-annual-and-perennial}{%
\subsubsection{Comparison of drought frequency between annual and
perennial}\label{comparison-of-drought-frequency-between-annual-and-perennial}}

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. The relationship between
drought frequencies across each taxon's range and life habitat (annual
or perennial) was evaluated using Firth's penalized-likelihood logistic
regression and phylogenetic logistic regression.

@@

Out of 8670 \emph{Heliophila} GBIF recrods, 6634 were for species with
reported life history (Mummenhoff et al., 2005), 3653 had geospatial
data, 3627 did not have geospatial issues, 3460 were located on pixels
classified as land having drought measurements, 3457 were located in
Africa, 3162 were not duplicated. After all filtering steps, 2192
records for 42 species (Figure \ref{fig:mapsboxplots}, Table
\ref{tab:speciesmeanstable}) passed.

Drought frequency was signicantly different between annual and perennial
species.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/maps_boxplots} \caption{Locations of annual and perennial Heliophila. Drought
frequency during the winter, spring, summer and fall. Drought
frequencies during each season observed at the collection sites of
Heliophila records.}\label{fig:mapsboxplots}
\end{figure}
